### 2022-03-04 bools/bools.tex
more

\looseness=-1 Terms are defined inductively as follows. Free ($X : \tau$)  and bound ($x : \tau$) variables   are
terms of type $\tau$. If $\cst{f} : \forallty{\tuplen{\alpha}} \tau$ is in
$\Sigma$ and $\tuplen{\upsilon}$ is a tuple of types, called type arguments, then
$\cst{f}\typeargs{\tuplen{\upsilon}}$ (written as $\cst{f}$ if $n=0$, or if type
arguments can be inferred from the context) is a term of type $\tau \{
\tuplen{\alpha} \mapsto \tuplen{\upsilon} \}$, called constant. If $x$ is a bound variable
of type $\tau$ and $s$ is a term of type $\upsilon$ then $\lamx{s}$ is a term of type
$\tau \rightarrow \upsilon$. If $s$ and $t$ are of type $\tau \rightarrow \upsilon$ and
$\tau$, respectively, then $s \, t$ is a term of type $\upsilon$. We call terms of
Boolean type ($o$) \emph{formulas} and denote them by $f,g,h, \ldots$; we use
$p,q,r, \ldots$ for free variables whose result type is $o$ and
$\cst{p},\cst{q},\cst{r}$ for constants with the same result type.
%
Unless stated otherwise, we view terms as
$\alpha\beta\eta$-equivalence classes, with $\eta$-long $\beta$-reduced form as
the representative.

@@

>>>>

\looseness=-1 Terms are defined inductively as follows. Free ($X : \tau$)  and bound ($x : \tau$) variables   are
terms of type $\tau$. If $\cst{f} : \forallty{\tuplen{\alpha}} \tau$ is in
$\Sigma$ and $\tuplen{\upsilon}$ is a tuple of types, called type arguments, then
$\cst{f}\typeargs{\tuplen{\upsilon}}$ (written as $\cst{f}$ if $n=0$, or if type
arguments can be inferred from the context) is a term of type $\tau \{
\tuplen{\alpha} \mapsto \tuplen{\upsilon} \}$, called constant. If $x$ is a bound variable
of type $\tau$ and $s$ is a term of type $\upsilon$ then $\lamx{s}$ is a term of type
$\tau \rightarrow \upsilon$. If $s$ and $t$ are of type $\tau \rightarrow \upsilon$ and
$\tau$, respectively, then $s \, t$ is a term of type $\upsilon$. We call terms of
Boolean type ($o$) \emph{formulas} and denote them by $f,g,h, \ldots$; we use
$p,q,r, \ldots$ for free variables whose result type is $o$ and
$\cst{p},\cst{q},\cst{r}$ for constants with the same result type.
%
Formulas whose top-level symbol is not logical are called \emph{atoms}.
Unless stated otherwise, we view terms as
$\alpha\beta\eta$-equivalence classes, with $\eta$-long $\beta$-reduced form as
the representative.

@@

\section{The Native Approach} 
\label{sect:bool:native}

\looseness=-1
Zipperposition already had some support for Boolean reasoning before we
started extending the calculus of Bentkamp et al. In this section, we first
describe the internals of Zipperposition responsible for reasoning with
Booleans. We continue by describing \NumberOK{15} rules that we have
implemented. For ease of presentation we divide them in three categories.
\subsection{Support for Booleans in Zipperposition}
\label{subsect:bool:zip-bools}

As mentioned in Chapter~\ref{ch:unif}, Zipperposition is an open source prover
written in OCaml. From its inception, it was designed as a prover that supports
easy extension of its base superposition calculus to various theories, including
arithmetic, induction and limited support for higher-order logic
\cite{sc-15-simon-phd}.

In Zipperposition, applications are represented in flattened, spine notation. In
addition, Zipperposition uses associativity of $\iand$ and $\ior$ to flatten out
the nested applications of these symbols. For example, terms $\cst{p} \iand
(\cst{q} \iand \cst{r})$ and $(\cst{p} \iand \cst{q}) \iand \cst{r}$ are
internally stored as $\iand \, \cst{p} \, \cst{q} \, \cst{r}$. 
Zipperposition's support for $\lambda$-terms is used to represent quantified nested
Booleans: Formulas $\iforall x. \, \, f$ and $\iexists x. \, f$ are represented as
$\iforall \, (\lamx{f})$ and $\iexists \, (\lamx{f})$. After clausification of the
input problem, no nested Booleans will be modified or renamed using fresh
predicate symbols.

The version of Zipperposition preceding our modifications distinguished between non\-equational
and equational literals. Following E \cite{scv-19-e23}, we
modified Zipperposition to represent all literals equationally: a nonequational
literal $f$ is stored as $f \eq \itrue$, whereas $\neg f$ is stored as $f
\noteq \itrue$. Equations of the form $f \eq \ifalse $ and $f \noteq \ifalse$ are
transformed into $f \noteq \itrue$ and $f \eq \itrue$, respectively.

\subsection{Core Rules}
\label{subsect:bool:core}

\looseness=-1
Kotelnikov et al.\ \cite{kotelnikov-15-fool}, to the best of our
knowledge, pioneered the approach of extending a first-order superposition prover to support
nested Booleans. They call effects of including the axiom $\iforall
p. \; p \eq \itrue \ior p \ieq \ifalse$ a  ``recipe for disaster''. To combat the
explosive behavior of the axiom, they imposed the following two requirements to
the simplification order $\succ$: $\itrue \succ \ifalse$ and
$\itrue$ and $\ifalse$ are two smallest ground terms with respect to $\succ$. If
these requirements are met, there is no self-paramodulation of the clause
and only paramodulation possible is from literal $p \eq \itrue$ of the mentioned axiom
into a Boolean subterm of another clause. Finally, Kotelnikov et al.\ replace
the axiom with the inference rule \emph{FOOL Paramodulation} (\infname{FP}):
%
$$ \namedinference{FP}{C[f]}{C[\itrue] \vee f \eq \ifalse} $$
%
\looseness=-1
where $f$ is a nested non-variable Boolean subterm of clause $C$, different from
$\itrue$ and $\ifalse$. In addition, they translate the initial problem containing nested
Booleans  to first-order logic without interpreted Booleans; thus, symbols $\itrue$ and $\ifalse$, and type $o$
correspond to proxy symbols and types introduced during the translation.

We created two rules that are syntactically similar to \infname{FP} but
are adapted for higher-order logic with one key distinction -- we do not perform any translation:
%
\begin{align*}
  & \namedinference{Cases}{C[f]}{C[\ifalse] \vee f \eq \itrue}
  && \namedsimp{CasesSimp}{C[f]}{C[\ifalse] \vee f \eq \itrue \quad C[\itrue] \vee f \noteq \itrue}
\end{align*}
%
\looseness=-1
Clearly, the prover that uses the rules should not include them both at the same time.
In addition, since
literals $f \eq \ifalse$ are represented as negative equations $f \noteq \itrue$, which cannot be used to paramodulate from,
we change the first requirement on the order to $\ifalse \succ
\itrue$.

\newcommand{\eqneq}{\mathrel{\dot{\eq}}}
These two rules hoist Boolean subterms $f$ to the literal level; therefore,
some results of \infname{Cases} and \infname{CasesSimp} will have literals of the form $f \eq \itrue$ (or
$f \noteq \itrue$) where $f$ is not an atom. This introduces the need for the rule
called eager clausification (\infname{EC}):
%
$$ \namedsimp{EC}{C}{D_1 \; \cdots \; D_m} $$
%
We say that a clause is \emph{standard} if all of its literals are of the form $s \eqneq t$,
where $s$ and $t$ are not Booleans or of the form $f \eqneq \itrue$, where the head of $f$
is not a logical symbol and $\eqneq$ denotes $\eq$ or $\not\eq$. The rule \infname{EC}
is applicable if clause $C = L_1 \llor
\cdots \llor L_n$ is not standard.
The resulting clauses $\tuple{D}{m}$ represent
the result of clausification of the formula $\iforall\,\tuple{x}{}. \; L_1 \ior
\cdots \ior L_n$ where $\tuple{x}{}$ are all free variables of $C$.
Using Boolean extensionality, Zipperposition's clausification
algorithm treats Boolean equality as equivalence (i.e., it replaces
$\mathop{\eq}\typeargs{o}$ with $\iimplies$).

\looseness=-1
An advantage of leaving nested Booleans unmodified is that the prover will be able
to prove some problems containing them without using the prolific rules described
above. For example, given two clauses $\cst{f} \, (\cst{p} \, x
\iimplies \cst{p} \, y) \eq \cst{a}$ and $\cst{f} \, (\cst{p} \,
\cst{a} \iimplies \cst{p} \, \cst{b}) \noteq \cst{a}$, the empty clause can
easily be derived without the above rules. A disadvantage of this approach
is that the proving process will periodically be interrupted by expensive calls
to the clausifier.
\pagebreak[2]

\looseness=-1
If implemented naively, rules \infname{Cases} and \infname{CasesSimp} can result in many redundant clauses. Consider the following
example: let $\cst{p} : o \rightarrow o$, $\cst{a} : o$ and consider a clause
set containing $\cst{p} \, (\cst{p} \, (\cst{p} \, (\cst{p} \, \cst{a}))) \eq
\itrue$. Then, the clause $C = \cst{a} \eq \itrue \llor \cst{p} \, \ifalse \eq \itrue$ can be
derived in eight ways using the rules, depending on which nested Boolean subterm was
chosen for the inference. In general, if a clause has a subterm occurrence of the form $\cst{p}^n \, \cst{a}$,
where both $\cst{p}$ and $\cst{a}$ have result type $o$, the clause $\cst{a} \eq \itrue \llor \cst{p} \, \ifalse \eq \itrue$ can be derived in $2^{n-1}$ ways.
% Boolean subterms occurring in different
% eligible literals cause similar explosive behavior.
To combat these issues we
implemented pragmatic restrictions of the rule: only $f$ which is
the leftmost outermost (or innermost) eligible subterm will be considered. With
this modification $C$ can be derived in only one way. Furthermore,
some intermediate conclusions of the rules will not be derived, pruning the search space.

The clausification algorithm by Nonnengart and Weidenbach \cite{nw-01-small-cnf}
aggressively simplifies the input problem using well-known Boolean equivalences before clausifying it. For example,
the formula $\cst p\land\itrue$ will be replaced by $\cst p$. To simplify nested Booleans we implemented the rule
\pagebreak[2]
\[
\namedsimp{BoolSimp}{C[f\sigma]}{C[g\sigma]}
\]
where $f \longrightarrow g \in E$ runs over fixed set of rewrite rules $E$,
and $\sigma$ is any substitution. In the current implementation of Zipperposition, $E$ contains the rules
described by Nonnengart and Weidenbach \cite[Section~3]{nw-01-small-cnf}. This set contains
the rules describing how each logical symbol behaves when given arguments $\itrue$ or $\ifalse$: for example, 
it includes $(\itrue \iimplies p)~\longrightarrow~p $ and $(p \iimplies \itrue) \longrightarrow \itrue$. %full list visa report Sect 4.2
Leo-III implements a similar rule, called \textsf{simp} \cite[Section 4.2.1.]{as-18-phd}.

Our decision to represent negative atoms as negative equations was motivated by
the need to alter Zipperposition's earlier behavior as little as possible. 
Namely, negative atoms were not used as literals that can be used
to paramodulate from, and as such added to the laziness of the superposition calculus.
However, it might be useful to consider unit clauses of the form $f \noteq \itrue$
as $f \eq \ifalse$ to strengthen rewriting. To that end, we have introduced the following
rule:
%
$$ \namedsimp{BoolDemod}{f \noteq \itrue \qquad C[f\sigma]}{f \noteq \itrue \qquad C[\ifalse]} $$
